prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2022-06-07 inst/MyArticle/paper.Rmd
Acknowledgements update
conclusions

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model quantitative proteomics experiments. Various types of models are available (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)), and the contrast specification is explicit and consistent for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for adding new features, e.g., generalized linear models (_glm_'s) to model the presence or absence information of a protein, or robust linear models (_rlm_'s), in the future. R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical methods programming languages. The developed framework is flexible to integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms].

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the $pAUC$, and if the FDR is an unbiased estimate of the FDR. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice. First, estimate protein abundances from peptide abundances using a robust or nonparametric regression method. Second, use linear models because they show good performance in all categories. If the measurements are correlated, as for technical replicates, mixed effect models might work if the sample sizes are large; if not, aggregate the replicates and fit a linear model instead. Third, if you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates. If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the difference. Last but not least, do not impute missing observation but rather statistically model missingness to estimate parameters, i.e., group differences.

In summary, `r BiocStyle::Githubpkg("fgcz/prolfqua")` is an easy-to-use and feature-rich R package to analyze quantitative mass spectrometric data with simple and complex experimental designs. It also can generate conclusive reports, and benchmark MS software and statistical methods. Furthermore, with very little adaptations this R package can also be adapted to analyze different quantitative proteomics data (e.g. labeling based TMT-, PRM- DIA-data or even other omics scale datasets such as microarray data).
We provide documentation, in vignette format, at the website https://github.com/fgcz/prolfqua/. This document was created using Rmarkdown. All the code needed to replicate the document or the benchmark results is available at : https://github.com/wolski/prolfquabenchmark.

@@

The authors thank the technology platform fund (TPF) of the University of Zurich.

>>>>

`r BiocStyle::Githubpkg("fgcz/prolfqua")` allows for considerable flexibility to model quantitative proteomics experiments. Various types of models are available (see Figure \@ref(fig:ContrastUML) and Table \@ref(tab:prolfquaModels)), and the contrast specification is explicit and consistent for all models. The modular design of `r BiocStyle::Githubpkg("fgcz/prolfqua")`, allows for adding new features, e.g., generalized linear models (_glm_'s) to model the presence or absence information of a protein, or robust linear models (_rlm_'s), in the future. R's formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with `r BiocStyle::Githubpkg("fgcz/prolfqua")` in other statistical methods programming languages. However, the developed framework is flexible enough to in the future integrate other modeling methods, e.g., the probabilistic dropout model [@bioxrvproDA2020] or accurate variance estimation [@zhu2020deqms]. Hence, `r BiocStyle::Githubpkg("fgcz/prolfqua")` enables you to call various methods and makes selecting the best differential expression analysis algorithm for your problem easy.

When comparing statistical modeling methods for the differential expression analysis, we assessed performance measures such as the number of estimated contrasts, the $pAUC$, and if the FDR is an unbiased estimate of the FDR. It is relevant that an analysis pipeline shows good performance in all these categories. Leveraging these computer experiments, we can provide the following advice: i) estimate protein abundances from peptide abundances using a robust or nonparametric regression method; ii) use linear models because they show good performance in all categories; iii) if the measurements are correlated, as for technical replicates, mixed effect models might work if the sample sizes are large; if not, aggregate the replicates and fit a linear model instead; iv) if you use fixed-effect linear models, apply variance moderation to improve the $t$-statistics and $p$-value estimates; v) If you want to sort your protein lists to perform gene set enrichment analysis, use the $t$-statistic instead of the difference; vi) do not impute missing observation but statistically model missingness to estimate parameters, i.e., group differences. Finally, the differential expression analysis result obtained with prolfqua are comparable to or better than when using other differential expression analysis tools.

In summary, `r BiocStyle::Githubpkg("fgcz/prolfqua")` is an easy-to-use and feature-rich R package to analyze quantitative mass spectrometric data with simple or complex experimental designs. It also can generate conclusive reports and to benchmark MS software and statistical methods. Furthermore, with minimal adaptations, this R package can analyze different quantitative proteomics data (e.g., labeling-based TMT-, PRM- DIA-data).
We provide documentation, in vignette format, at the website https://github.com/fgcz/prolfqua/. This document was created using Rmarkdown. All the code needed to replicate the document or the benchmark results is available at: https://github.com/wolski/prolfquabenchmark.

@@

The authors thank the technology platform fund (TPF) of the University of Zurich and all FGCZ proteomics colleagues for fruitful discussions.
